Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
reorg.

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point

@@

%==============================================================
\begin{figure*}[t]
\renewcommand{\arraystretch}{1.5}
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
$\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha
\mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne
\emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
\mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
\begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{\epath} \mapsto
\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{\epath}$ is a new e-path
  terminating with $\psi$}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) =
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
\ldots \mathit{df_k}$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

\section{Computing liveness and its encoding as DFA}\label{sec:computing}
Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
and  \mycomment{\color  {red}  proved  it  correct  with  reference  to  a  {\em
    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
together described a liveness set for each program point in a function
in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
describe how to obtain demand transformers \Lfonly\ from the rule {\sc
  live-define} and how to compute the specific demand $\sigma$ on each
function.
%%
To  answer these  questions, it  is convenient  to cast  the equations
arising  out of  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
grammar.   To do  so, we  need  to modify  the rules  themselves to  a
different form.

@@

\begin{table*}[ht!]
\caption{Comparing LGC times for different strategies}
\label{tab:compare-live-clo-reach-clo}
\centering
\begin{tabular}
{| l | r | r | r | r |}
\hline
Prog Name & Heap size & Num GC & Time taken using reachability for closure  & Time taken using liveness for closure \\ \hline
\verb@nqueens (reduced) @ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
\end{tabular}
\end{table*}

\begin{table*}[ht!]
\caption{Liveness Analysis and Garbage Collection statistics}
\label{tab:exp-results}
\centering
 \begin{tabular}{|l||c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c|} \hline
   \multicolumn{1}{|r||}{Program} & \verb@lambda@ & \verb@nperm@ & \verb@treejoin@ & \verb@lcss@ & \verb@sudoku@ & \verb@fibheap@ & \verb@nqueens@ & \verb@knightstour@ & \verb@gc_bench@ \\ \hline\hline
   \#CFG Nonterminals            & 1078          & 794          & 709             & 839         & 1758          & 780            & 529            & 659                & 417             \\ \hline
   \#CFG Rules                   & 1823          & 1038         & 1610            & 1752        & 2509          & 1358           & 748            & 909                & 486             \\  \hline 
   \#DFA States                  & 3312          & 2055         & 2332            & 2340        & 5672          & 2254           & 1229           & 1742               & 729             \\\hline
   \#DFA Transitions             & 5792          & 3632         & 4235            & 4261        & 10857         & 4076           & 2078           & 3101               & 1124            \\\hline
   DFA Generation Time (sec)      & 14.7          & 1.1          & 2690.5          & 13.3        & 798.5         & 49.8           & 0.5            & 5.1                & 0.1             \\\hline
 \end{tabular}\\
 (a) Data for Liveness Analysis \\[3pt]
\include{autotable}
\end{table*}

@@

\section{Experimental Evaluation}
\label{sec:experiments}
To   test  the   utility  of   our  method   we  have   implemented  a
garbage-collector for a simple scheme-like first order language having
non-strict semantics. It consists of an interpreter, liveness analyzer
and a  garbage-collector that can  optionally use  reachability or
liveness.  Our  benchmark consists  of programs  from nofib~\cite{nofib}  suite and
some external programs manually converted to ANF.

@@

Therefore we believe  that using reachability for closures  is a sweet
spot   in   terms   of   performance  for   liveness   based   garbage
collectors.  For the  remainder  of  the section  all  LGC results  we
discuss uses  a mixed-mode collector--liveness for  evaluated data and
reachability for closures.
%%
%% \comment{
%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
%% in the heap over time for RGC and LGC---here time is measured in terms
%% of the number of cons cells allocated. In addition, they also show the
%% number of  reachable cells and the  number of cells  that are actually
%% live\footnote{This   is  statically   approximated  by   our  liveness
%%   analysis}.  Since  the programs have  different memory requirements,
%% we have tuned the size of heap for each program to ensure a reasonable
%% number of  collections. An invocation  of RGC decreases the  number of
%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
%% invocation of LGC decreases the number  of heap cells to no lower than
%% the curve of live cells.

%% To construct the  reachable and live curves, we  record for every cell
%% its creation time (\CreateTime), its last use time (\UseTime), and the
%% earliest time  when the  cell becomes unreachable  and can  be garbage
%% collected (\CollTime).  For accurate recording of  \CollTime, we force
%% frequent invocations  of a reachability based collector  in a separate
%% run.     A    cell   is    live    at    time   \mbox{\exectime}    if
%% \mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
%%   \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
%%   \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.

%% The benchmark programs are drawn from the {\tt no-fib} suite and other
%% sources and have been  manually converted to  ANF.  All  graphs  except
%%  {\tt  fibheap}  show  fewer  garbage
%% collector  invocations for  LGC. The  number  of reachable  cells in
%% {\tt
%%   fibheap} increases  initially, comes  down before the  heap becomes
%% gets
%% full and then  continues to remains low.  This results in  a single
%% garbage
%% collection in both RGC and LGC.  The graphs also show the precision of
%% our liveness analysis.  For  all programs  except  {\tt nperms} and
%% {\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
%% manages to collect a good portion of the data that are not live.}
%%

\subsection{Results}
\newcommand{\hgt}{3cm}
\begin{figure*}[p]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}}
   \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
   & {\epsfig{file=lambda_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
\\ (a) & (b)
\end{tabular}%\vskip -10mm
 \caption{Memory usage.  Column (a) shows complete  usage while column
   (b) enlarges  a part  of the  usage.  The blue  and the  red curves
   indicate the number of cons cells  in the active semi-space for RGC
   and LGC  respectively.  The  black curve  represents the  number of
   reachable cells and  the grey curve represents the  number of cells
   that are  actually live (of  which liveness analysis does  a static
   approximation).   x-axis   is  the  time  measured   in  number  of
   cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
   number of cons-cells (scaled down by factor $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}

@@

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
average  number of  cells reclaimed  per GC,  average number  of cells
touched  per GC,  total  number of  dragged cells  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  GC is  higher (and  consequently the  number of  GCs is
lower)  for LGC  based collection.   LGC  also reduces  the number  of
dragged cells significantly. However, the cost  for LGC is that per GC
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within 5X of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthestic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs. All our benchmarks were executed
on  a desktop  with  64 bit  Ubuntu 14.04,  8  core Intel(R)  Core(TM)
i7-4770 CPU @ 3.40GHz, L2 cache of 8192 KB and 15922 MiB (16 GB) RAM.

>>>>

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
%==============================================================
\begin{figure*}[t]
\renewcommand{\arraystretch}{1.5}
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
$\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha
\mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne
\emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
\mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
\begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{\epath} \mapsto
\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{\epath}$ is a new e-path
  terminating with $\psi$}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) =
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
\ldots \mathit{df_k}$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

@@

\section{Computing liveness and its encoding as DFA}\label{sec:computing}
Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis%
  \mycomment{\color  {red}  and proved  it  correct  with  reference  to  a  {\em
    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
together described a liveness set for each program point in a function
in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
describe how to obtain demand transformers \Lfonly\ from the rule {\sc
  live-define} and how to compute the specific demand $\sigma$ on each
function.
%%
To  answer these  questions, it  is convenient  to cast  the equations
arising  out of  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
grammar.   To do  so, we  need  to modify  the rules  themselves to  a
different form.

@@

\begin{table}[ht!]
\caption{Time taken by LGC for different strategies.  The programs are
  modified (smaller) instances of original benchmarks.}
\label{tab:compare-live-clo-reach-clo}
\centering
\begin{tabular}
{| l | r | r | r |}
\hline
Program        & \#GC & \multicolumn{2}{c|}{Closure collected using} \\ \cline{3-4}
               &      & Reachability & Liveness                      \\ \hline
\verb@nqueens@ & 183  & 0.009 sec    & 0.228 sec                     \\\hline
\verb@lambda@  & 11   & 0.001 sec    & 4.992 sec                     \\ \hline
\verb@lcss@    & 1    & 0.000 sec    & 34.668 sec                    \\ \hline
\end{tabular}
\end{table}

@@

\section{Experimental Evaluation}
\label{sec:experiments}
To   test  the   utility  of   our  method   we  have   implemented  a
garbage-collector for a simple scheme-like first order language having
non-strict semantics. It consists of an interpreter, liveness analyzer
and  a  garbage-collector  that  can optionally  use  reachability  or
liveness.  Our benchmark consists  of programs from nofib~\cite{nofib}
suite and some  external programs manually converted to  ANF.  All our
benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
Ubuntu 14.04.

@@

Therefore we believe  that using reachability for closures  is a sweet
spot   in   terms   of   performance  for   liveness   based   garbage
collectors.  For the  remainder  of  the section  all  LGC results  we
discuss uses  a mixed-mode collector--liveness for  evaluated data and
reachability for closures.

\begin{table*}[ht!]
\caption{Statistics for liveness analysis and garbage collection}
\label{tab:exp-results}
\centering
 \begin{tabular}{|l||c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c|} \hline
   \multicolumn{1}{|r||}{Program} & \verb@lambda@ & \verb@nperm@ & \verb@treejoin@ & \verb@lcss@ & \verb@sudoku@ & \verb@fibheap@ & \verb@nqueens@ & \verb@knightstour@ & \verb@gc_bench@ \\ \hline\hline
   \#CFG Nonterminals            & 1078          & 794          & 709             & 839         & 1758          & 780            & 529            & 659                & 417             \\ \hline
   \#CFG Rules                   & 1823          & 1038         & 1610            & 1752        & 2509          & 1358           & 748            & 909                & 486             \\  \hline 
   \#DFA States                  & 3312          & 2055         & 2332            & 2340        & 5672          & 2254           & 1229           & 1742               & 729             \\\hline
   \#DFA Transitions             & 5792          & 3632         & 4235            & 4261        & 10857         & 4076           & 2078           & 3101               & 1124            \\\hline
   DFA Generation Time (sec)      & 14.7          & 1.1          & 2690.5          & 13.3        & 798.5         & 49.8           & 0.5            & 5.1                & 0.1             \\\hline
 \end{tabular}\\
 (a) Data for Liveness Analysis \\[3pt]
\include{autotable}
\end{table*}

\subsection{Results}

@@

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
average  number of  cells reclaimed  per GC,  average number  of cells
touched  per GC,  total  number of  dragged cells  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  GC is  higher (and  consequently the  number of  GCs is
lower)  for LGC  based collection.   LGC  also reduces  the number  of
dragged cells significantly. However, the cost  for LGC is that per GC
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within 5X of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthestic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs.

\newcommand{\hgt}{3cm}
\begin{figure*}[p]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}}
   \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
   & {\epsfig{file=lambda_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
\\ (a) & (b)
\end{tabular}%\vskip -10mm
 \caption{Memory usage.  Column (a) shows complete  usage while column
   (b) enlarges  a part  of the  usage.  The blue  and the  red curves
   indicate the number of cons cells  in the active semi-space for RGC
   and LGC  respectively.  The  black curve  represents the  number of
   reachable cells and  the grey curve represents the  number of cells
   that are  actually live (of  which liveness analysis does  a static
   approximation).   x-axis   is  the  time  measured   in  number  of
   cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
   number of cons-cells (scaled down by factor $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}
